---
title: "The Impact of Storms and Severe Weather Events on the Population and Economic Health of the United States (1950 - 2011)"
output: html_document
---

###Synopsis  
This analysis details the impact of storms and severe weather events over a 61-year period (1950 to 2011) on the U.S. population, in terms of injuries and deaths, as well as the associated dollar damage to properties and crops. The analysis uses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database for the raw data and then implements 16 data processing steps to convert the raw data into the plots and tables presented in the **Results** and **Data Sets** sections below.  As part of the data processing steps, a large cleanup exercise was undertaken to convert the NOAA event type field (i.e. EVTYPE) to one of the forty-eight (48) consistent NOAA storm types.  As with any classification exercise, liberties were taken to adhere to the 48 storm types; however, any rational reclassification, made by independent means, to assign event types to storm types would not change the results of this analysis in any material way.

It is suprising to note Tornados accounted for more than 62% of the injuries/fatalities sustained during this time-period.  In fact, Tornados outpaced the combined other 47 storm types by a 1.7 magnitude.  Although Tornados caused a significant number of injuiries/fatalities, they ranked 3rd in terms of economic damage ($ damage to property and crops).  Moreover, the top four storm types (Floods, Hurricanes (Typhoons), Tornados, and Storm Surge/Tides) accounted for approximately 75% of the economic damage caused during the time period.

For illustrative purposes, tabular results are presented in their entirely within the **Data Sets** section, while plots are streamlined in the **Results** section to show the event types which contribute most to the question being addressed.

Lastly, this analysis defines population health as the summation of any injuries or deaths, while economic damage is the summation (in $) of the amount of destruction caused to both property and crops.

###Data Processing    
This section describes the 16 steps used to process the NOAA storm data.  

**1.  Clear the screen and load the required libraries**
```{r, Data Processing Step 1}
rm(list=ls())   #Clear all objects

require(dplyr)
require(stringr)
```
**2.  Set the working directory and file vectors**  
```{r, Data Processing Step 2}
working.dir <- "C://Education//Reproducible Research//Homework//Peer Assessment 2"
setwd(working.dir)

input.dir <- ".\\input\\"
output.dir <- ".\\output\\"
script.dir <- ".\\scripts\\"
input.storm.data <- "repdata-data-StormData.csv"
input.storm.ref.data <- "storm-reference-data.csv"
```
**3.  Load the storm data**  
```{r, cache = TRUE}
storm.data <- read.csv(paste(input.dir, input.storm.data, sep="\\"), stringsAsFactors = FALSE, strip.white = TRUE)
storm.ref.data <- read.csv(paste(input.dir, input.storm.ref.data, sep="\\"), stringsAsFactors = FALSE, strip.white = TRUE)
```
**4.  Subset storm.data where pop.health > 0 | PROPDMG > 0 | CROPDMG > 0**  
```{r, cache = TRUE}
storm.data.step.1 <- storm.data %.%
  mutate(pop.health = FATALITIES + INJURIES, 
         prop.dmg.exp = 0,
         prop.dmg.value = 0,
         crop.dmg.exp = 0,
         crop.dmg.value = 0,
         total.dmg.value = 0,
         event.type.id = as.integer(0)) %.%
  select(event.type.id, EVTYPE, FATALITIES, INJURIES, pop.health, PROPDMG, PROPDMGEXP, prop.dmg.exp, prop.dmg.value, CROPDMG, CROPDMGEXP, crop.dmg.exp, crop.dmg.value, total.dmg.value) %.%
  filter(pop.health > 0 | PROPDMG > 0 | CROPDMG > 0)
```

**5.  Clean-up the subsetted storm data**  
```{r, Data Processing Step 5}
storm.data.step.1$PROPDMGEXP = toupper(storm.data.step.1$PROPDMGEXP)
storm.data.step.1$CROPDMGEXP = toupper(storm.data.step.1$CROPDMGEXP)
storm.data.step.1$EVTYPE <- str_trim(storm.data.step.1$EVTYPE)
```

**6.  Set the PROPDMGEXP and CROPDMGEXP variables**  
```{r, Data Processing Step 6}
storm.data.step.1[storm.data.step.1$PROPDMGEXP == "1", c("prop.dmg.exp")] = 10
storm.data.step.1[storm.data.step.1$PROPDMGEXP %in% c("H", "2"), c("prop.dmg.exp")] = 100
storm.data.step.1[storm.data.step.1$PROPDMGEXP %in% c("K", "3"), c("prop.dmg.exp")] = 1000
storm.data.step.1[storm.data.step.1$PROPDMGEXP == "4", c("prop.dmg.exp")] = 10000
storm.data.step.1[storm.data.step.1$PROPDMGEXP == "5", c("prop.dmg.exp")] = 100000
storm.data.step.1[storm.data.step.1$PROPDMGEXP %in% c("M", "6"), c("prop.dmg.exp")] = 1000000
storm.data.step.1[storm.data.step.1$PROPDMGEXP == "7", c("prop.dmg.exp")] = 10000000
storm.data.step.1[storm.data.step.1$PROPDMGEXP == "8", c("prop.dmg.exp")] = 100000000
storm.data.step.1[storm.data.step.1$PROPDMGEXP %in% c("B", "9"), c("prop.dmg.exp")] = 1000000000

storm.data.step.1[storm.data.step.1$CROPDMGEXP == "1", c("crop.dmg.exp")] = 10
storm.data.step.1[storm.data.step.1$CROPDMGEXP %in% c("H", "2"), c("crop.dmg.exp")] = 100
storm.data.step.1[storm.data.step.1$CROPDMGEXP %in% c("K", "3"), c("crop.dmg.exp")] = 1000
storm.data.step.1[storm.data.step.1$CROPDMGEXP == "4", c("crop.dmg.exp")] = 10000
storm.data.step.1[storm.data.step.1$CROPDMGEXP == "5", c("crop.dmg.exp")] = 100000
storm.data.step.1[storm.data.step.1$CROPDMGEXP %in% c("M", "6"), c("crop.dmg.exp")] = 1000000
storm.data.step.1[storm.data.step.1$CROPDMGEXP == "7", c("crop.dmg.exp")] = 10000000
storm.data.step.1[storm.data.step.1$CROPDMGEXP == "8", c("crop.dmg.exp")] = 100000000
storm.data.step.1[storm.data.step.1$CROPDMGEXP %in% c("B", "9"), c("crop.dmg.exp")] = 1000000000
```

**7.  Set the Damage Value variables**  
```{r, Data Processing Step 7}
storm.data.step.1$prop.dmg.value = as.numeric(storm.data.step.1$PROPDMG) * as.numeric(storm.data.step.1$prop.dmg.exp)
storm.data.step.1$crop.dmg.value = as.numeric(storm.data.step.1$CROPDMG) * as.numeric(storm.data.step.1$crop.dmg.exp)
storm.data.step.1$total.dmg.value = storm.data.step.1$prop.dmg.value + storm.data.step.1$crop.dmg.value
```

**8.  Summarize the Population Health and Damage Values by the Event Type**  
```{r, Data Processing Step 8}
storm.data.step.2 <- storm.data.step.1 %.%  
  group_by(EVTYPE) %.%
  summarise(event.type.id = sum(event.type.id), sum.pop.health = sum(pop.health), sum.prop.dmg.value = sum(prop.dmg.value), sum.crop.dmg.value = sum(crop.dmg.value), sum.total.dmg.value = sum(total.dmg.value)) %.%
  select(event.type.id, EVTYPE, sum.pop.health, sum.prop.dmg.value, sum.crop.dmg.value, sum.total.dmg.value)
```

**9.  Cast ids to their representative data types**  
```{r, Data Processing Step 9}
storm.data.step.2$id = as.numeric(rownames(storm.data.step.2))
storm.data.step.2$event.type.id = as.integer(storm.data.step.2$event.type.id)
```

**10.  Assign the EVTYPE field to one of the 48 NOAA storm types**  
```{r, Data Processing Step 10}
storm.data.step.2[grep("LIGHT", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 29
storm.data.step.2[grep("LIGNTNING", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 29
storm.data.step.2[grep("SLIDE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 6
storm.data.step.2[grep("Landslump", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 6
storm.data.step.2[grep("WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 38
storm.data.step.2[grep("WATER", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 45
storm.data.step.2[grep("TORNADO", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 40
storm.data.step.2[grep("TORNDAO", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 40
storm.data.step.2[grep("LANDSPOUT", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 40
storm.data.step.2[grep("GUSTNADO", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 40
storm.data.step.2[grep("TSUNAMI", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 43
storm.data.step.2[grep("TROPICAL STORM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 42
storm.data.step.2[grep("TROPICAL DEPRESSION", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 41
storm.data.step.2[grep("THUNDER", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 39
storm.data.step.2[grep("BLIZZARD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 3
storm.data.step.2[grep("FREEZE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 17
storm.data.step.2[grep("FREEZI", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 17
storm.data.step.2[grep("FROST", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 17
storm.data.step.2[grep("HYPERTHERMIA/EXPOSURE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 17
storm.data.step.2[grep("HYPOTHERMIA", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 17
storm.data.step.2[grep("LOW TEMPERATURE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 17
storm.data.step.2[grep("ASTRONOMICAL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 1
storm.data.step.2[grep("AVALANC", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 2
storm.data.step.2[grep("HURRIC", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 25
storm.data.step.2[grep("TYPHOON", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 25
storm.data.step.2[grep("ICE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 26
storm.data.step.2[grep("ICY ROADS", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 26
storm.data.step.2[grep("GLAZE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 25
storm.data.step.2[grep("RAIN", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 21
storm.data.step.2[grep("HEAVY MIX", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 21
storm.data.step.2[grep("HEAVY SEA", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 21
storm.data.step.2[grep("HEAVY SHOWER", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 21
storm.data.step.2[grep("HEAVY PRECIPITATION", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 21

storm.data.step.2[grep("HEAVY SURF", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("HIGH SURF", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("ROUGH SURF", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("ROUGH SEA", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("HIGH SWELLS", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("HEAVY SWELLS", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("HIGH TIDE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("HIGH WAVES", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23
storm.data.step.2[grep("HIGH SEA", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 23

storm.data.step.2[grep("SLEET", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 36
storm.data.step.2[grep("SNOW", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 22
storm.data.step.2[grep("FLOOD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 15
storm.data.step.2[grep("URBAN", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 15
storm.data.step.2[grep("WETNESS", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 15
storm.data.step.2[grep("FLASH FLOOD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 14
storm.data.step.2[grep("COASTAL FLOOD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 4
storm.data.step.2[grep("LAKE FLOOD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 28
storm.data.step.2[grep("LAKESHORE FLOOD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 28
storm.data.step.2[grep("FOG", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 7
storm.data.step.2[grep("DENSE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 7
storm.data.step.2[grep("FREEZING FOG", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 16
storm.data.step.2[grep("DUST", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 10
storm.data.step.2[grep("DUST STORM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 11
storm.data.step.2[grep("WINTER STORM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 47
storm.data.step.2[grep("WINTER WEATHER", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 48
storm.data.step.2[grep("WINTRY MIX", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 48
storm.data.step.2[grep("MIXED PRECIP", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 48
storm.data.step.2[grep("FIRE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 46
storm.data.step.2[grep("COLD", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 5
storm.data.step.2[grep("COOL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 5
storm.data.step.2[grep("WIND CHILL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 5
storm.data.step.2[grep("WINDCHILL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 5
storm.data.step.2[grep("HEAT", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 20
storm.data.step.2[grep("UNSEASONABLY WARM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 20
storm.data.step.2[grep("WARM WEATHER", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 20
storm.data.step.2[grep("DROUGHT", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 9
storm.data.step.2[grep("Beach Erosion", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("COASTAL EROSION", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("Erosion/Cstl Flood", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("COASTAL STORM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("COASTALSTORM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("COASTAL SURGE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("HAZARDOUS SURF", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("ROGUE WAVE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37
storm.data.step.2[grep("STORM SURGE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 37

storm.data.step.2[grep("DAM BREAK", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 14
storm.data.step.2[grep("DOWNBURST", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 14
storm.data.step.2[grep("HAIL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 19
storm.data.step.2[grep("MARINE HAIL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 30
storm.data.step.2[grep("RIP CURRENT", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 34
storm.data.step.2[grep("DROWNING", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 34
storm.data.step.2[grep("BURST", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 39
storm.data.step.2[grep("FUNNEL", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 18
storm.data.step.2[grep("SEICHE", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 35

storm.data.step.2[grep("HIGH WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 24
storm.data.step.2[grep("STRONG WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 38
storm.data.step.2[grep("MARINE HIGH WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 31
storm.data.step.2[grep("MARINE STRONG WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 32
storm.data.step.2[grep("MARINE THUNDERSTORM WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 33
storm.data.step.2[grep("THUNDERSTORM WIND", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 39
storm.data.step.2[grep("TSTM", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 39
storm.data.step.2[grep("VOLCANIC ASH", storm.data.step.2$EVTYPE, ignore.case=T), c("event.type.id")] <- 44
```

**11.  Cast event.type.ids**  
```{r, Data Processing Step 11}
storm.ref.data$event.type.id = as.integer(storm.ref.data$event.type.id)
storm.data.step.2$event.type.id = as.integer(storm.data.step.2$event.type.id)
```

**12.  Merge the storm data with the associated storm data hierarchy**  
```{r, Data Processing Step 12}
storm.data.step.2 <- left_join(storm.data.step.2, storm.ref.data)
```

**13.  Summarize the merged storm data by the event group/event name**  
```{r, Data Processing Step 13}
storm.data.step.3 <- storm.data.step.2 %.%  
  group_by(event.type.group, event.name) %.%
  summarise(sum.pop.health = sum(sum.pop.health), sum.prop.dmg.value = sum(sum.prop.dmg.value), sum.crop.dmg.value = sum(sum.crop.dmg.value), sum.total.dmg.value = sum(sum.total.dmg.value)) %.%
  select(event.type.group, event.name, sum.pop.health, sum.prop.dmg.value, sum.crop.dmg.value, sum.total.dmg.value)
```

**14.  Prepare the Population Health's Vectors and Data.Frames**  
```{r, Data Processing Step 14}
sum.pop.health <- sum(storm.data.step.3$sum.pop.health) 
max.pop.health <- max(storm.data.step.3$sum.pop.health)
diff.pop.health <- sum.pop.health - max.pop.health
per.diff.pop.health <- round((max.pop.health/sum.pop.health) * 100, 1)
pop.health.magnitude <- round(max.pop.health/diff.pop.health, 1)
max.pop.health.event <- as.character(subset(storm.data.step.3, sum.pop.health == max.pop.health, select = c(event.name)))
other.pop.health.event <- "All other event types combined"
names.argument.pop.health <- as.vector(c(max.pop.health.event, other.pop.health.event))
bar.plot.data <- as.data.frame(rbind(max.pop.health, diff.pop.health))
rownames(bar.plot.data) <- NULL
colnames(bar.plot.data) <- c("pop.health")
bar.plot.data$pop.health <- as.numeric(bar.plot.data$pop.health)
```

**15.  Prepare the Economic Consequence Vectors and Data.Frames**  
```{r, Data Processing Step 15}
econ.plot.data <- storm.data.step.3[ order(-storm.data.step.3[,6]), c(2,6)]
econ.top.4 <- sum(econ.plot.data[1:4, c(2)])
econ.all <- sum(econ.plot.data[ c(2)])
econ.pct <- round((econ.top.4/econ.all)*100, 1)
```

**16.  Create a human-friendly view of the storm data**  
```{r, Data Processing Step 16}
storm.data.step.4 <- subset(storm.data.step.3, is.na(event.name) == F, c("event.name", "sum.pop.health", "sum.total.dmg.value")) #With non-user friendly names
#storm.data.step.5 <- storm.data.step.4 
#colnames(storm.data.step.5) <- c("Event Name", "Number of Injuries and Fatalities", "Property and Crop Damage ($)") #With user friendly names
```

###Results    
**1.  Across the United States, which types of events are most harmful with respect to population health?**  
The NOAA storm data shows Tornados accounted for `r max.pop.health` out of the `r sum.pop.health` (`r per.diff.pop.health `%) injuries/deaths for the period.  In fact, Tornados caused approximately `r pop.health.magnitude` times more injuries/fatalities than the other 47 storm types combined.  This is highlighted by the plot below.  

Please see the **Data Sets** section for a listing of injuries/deaths by storm type.

```{r, Results Step 2}
x.label <- "Weather Event"
y.label <- "Number of Injuries/Fatalities"
y.limit <- c(0, 100000)
main.title <- "Total Number of U.S. Weather Related Injuries/Deaths (1950 - 2011)"
barplot(bar.plot.data$pop.health, names.arg = names.argument.pop.health, xlab = x.label, ylab = y.label, ylim = y.limit, main = main.title)
```

**2.  Across the United States, which types of events have the greatest economic consequences?**  
The NOAA storm data shows Floods, Hurricanes (Typhoons), Tornados, and Storm Surge/Tides accounted for $`r econ.top.4` out of the $`r econ.all` (`r econ.pct `%) property and/or crop damage for the period.  This is highlighted by the plot below.  

Please see the **Data Sets** section for a listing of $ amount of property and/or crop damage by storm type.
```{r, Results Step 3}
x.label <- "Weather Event"
y.label <- "$ Damage (Property/Crop)"
y.limit <- c(0, 200000000000)
main.title <- "Total $ of U.S. Weather Related Property/Crop Damage (1950 - 2011)"
df <- storm.data.step.3[ order(-storm.data.step.3[,6]), c(2,6)]
barplot(df[1:4, c(2)], names.arg = df[1:4, c(1)], xlab = x.label, ylab = y.label, ylim = y.limit, main = main.title)
```

###Data Sets 
**1.  Number of Injuries/Fatalities By Storm Type (1950 - 2011)**
```{r, Data Sets 1}
storm.data.step.6 <- subset(storm.data.step.4, sum.pop.health > 0, c(event.name, sum.pop.health))
colnames(storm.data.step.6) <- c("Event Name", "Number of Injuries and Fatalities")
storm.data.step.6[ order(-storm.data.step.6[,2]), c("Event Name", "Number of Injuries and Fatalities")]
```

**2.  $ Amount of Property and/or Crop Damage By Storm Type (1950 - 2011)**
```{r, Data Sets 2}
storm.data.step.6 <- subset(storm.data.step.4, sum.total.dmg.value > 0, c(event.name, sum.total.dmg.value))
colnames(storm.data.step.6) <- c("Event Name",  "Property and Crop Damage ($)")
storm.data.step.6[ order(-storm.data.step.6[,2]), c("Event Name",  "Property and Crop Damage ($)")]
```
